Why Real-World Subgraph Isomorphism is Easy, \\ and What This Means for Graph Databases

### 2017-07-22 paper/paper.tex
Christine's comments

\begin{abstract}
    Despite subgraph isomorphism being \NP-complete, modern algorithms can operate comfortably on
    most real-world graphs with thousands of vertices. However, they cannot handle arbitrary instances of
    this size. We show how to generate ``really hard'' random instances for subgraph isomorphism
    problems, which are computationally challenging with only a couple of hundred vertices in the
    target, and only twenty to thirty pattern vertices. For the non-induced version of the problem,
    these instances lie on a satisfiable / unsatisfiable phase transition, whose location we can
    predict; for the induced variant, much richer behaviour is observed, and constrainedness gives a
    better measure of difficulty than does proximity to a phase transition. In both cases, we can
    use this to understand and improve variable and value ordering heuristics.

@@

However, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances involving hundreds
or thousands of vertices. Experimental evaluations of subgraph isomorphism algorithms are usually
performed using randomly generated graph pairs, sometimes together with a mix of real-world graphs,
and graphs that encode biochemistry and computer vision problems. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. The random instances
used typically come from common datasets
\citep{DBLP:journals/prl/SantoFSV03,DBLP:journals/constraints/ZampelliDS10}, which were generated by
taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh)
graph and permuting the vertices. Such instances are guaranteed to be
satisfiable---\citet{o:AntonO09} exploited this property to create large sets of random satisfiable
boolean satisfiability instances.  This is the most common approach to generating random subgraph
isomorphism instances, meaning existing benchmark suites contain relatively few non-trivial
unsatisfiable instances (although a few of the patterns in the instances by
\citeauthor{DBLP:journals/constraints/ZampelliDS10} have had extra edges added, making them
unsatisfiable). Also, the satisfiable instances tend to be computationally fairly easy, with most of
the difficulty being in dealing with the size of the model. This has lead to bias in algorithm
design, to the extent that some proposed techniques, such as those of
\citet{DBLP:conf/sls/BattitiM07}, will \emph{only} work on satisfiable instances.

The first contribution of this paper is to present and evaluate new methods for creating random
pattern~/ target pairs.  The method we introduce generates both satisfiable and unsatisfiable
instances, and can produce computationally challenging instances with only a few tens of vertices in
the pattern, and 150 vertices in the target. This is not entirely straightforward---the lack of
unsatisfiable instances for testing purposes cannot be addressed simply by taking a pattern graph
from one of the existing random suites with the ``wrong'' target graph, as this tends to give either
a trivially unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the
case that a relatively small random graph is unlikely to appear in a larger random graph.)

This work builds upon the phase transition phenomena observed in satisfiability and graph colouring
problems first described by \citet{DBLP:conf/ijcai/CheesemanKT91} and
\citet{DBLP:conf/aaai/MitchellSL92}. For subgraph isomorphism we identify three relevant control
parameters: we can independently alter the edge probability of the pattern graph, the edge
probability of the target graph, and the relative orders (number of vertices) of the pattern and
target graphs.  For non-induced isomorphisms, with the correct choice of parameters we see results
very similar to those observed with boolean satisfiability problems: there is a phase transition
(whose location we can predict) from satisfiable to unsatisfiable, and we see a solver-independent
complexity peak near this phase transition. Additionally, understanding this behaviour
helps us to select better variable- and value-ordering heuristics---this is the second contribution
of this paper.

@@

When labels on vertices are introduced, as is commonly seen in real-world applications and in graph
database systems, richer behaviour emerges, particularly when moving away from a uniform labelling
scheme, and we see that VF2 behaves much worse than the Glasgow and LAD algorithms in certain cases.
A close look shows that these cases \emph{should} be easy to solve.

@@

We look at some of the datasets commonly used to test graph database systems, which the literature
suggests are extremely hard to solve. Simple experiments show that the entire perceived difficulty
of each of these datasets comes down to the widespread use of especially poor subgraph isomorphism
algorithms. Our fourth contribution is to show that the use of forward-checking and the right
variable ordering heuristic, even without the more sophisticated recent advances in subgraph
algorithms, makes the entire graph database filter~/ verify paradigm unnecessary.  Finally, we
explain why filter~/ verify (and other recently proposed techniques) cannot be beneficial even on
more challenging instances except if an extremely poor choice of subgraph isomorphism algorithm is
made.

\section{Experimental Setup}

@@

Recall that the Glasgow and LAD solvers use backtracking search to build up an assignment of target
vertices (values) to pattern vertices (variables), but differ in terms of inference and ordering
heuristics.  The approach used by VF2 is similar, although the domains of variables are not stored
(in the style of conventional backtracking, rather than forward-checking), and so domain wipeouts
are not detected until an assignment is made. Both the Glasgow and LAD solvers use domain sizes to
guide search, whilst VF2 uses adjacency to previously-assigned vertices.

@@

@@

@@

In the first four rows of \cref{figure:induced} we repeat our experiments, finding induced
isomorphisms. With a pattern of order 10, we get two independent phase transitions: the bottom right
half of the plots resemble the non-induced results, and the top left half is close to a mirror
image. The central satisfiable region, which is away from either phase transition, is
computationally easy, but instances near the phase transition are hard.

@@

What about empirical hardness? As before, some instances on the phase transition are hard for all
solvers (although as the number of labels increases, the hardest instances become easier).
Instances far from the phase transition are easy, except for VF2, which occasionally finds
some of instances with larger numbers of labels very difficult. This behaviour occurs both on
satisfiable instances, and on the flawed unsatisfiable instances deep inside the ``satisfiable''
region.  It is interesting to observe that all such unsatisfiable instances that VF2 finds difficult
have a very small proof of unsatisfiability, with neither Glasgow nor LAD requiring more than one
hundred recursive calls.

@@

We plot the cumulative number of instances solved over time for these instances in
\cref{figure:skewed-cumulative}. Both Glasgow and LAD find all of these instances trivial, but VF2
finds many of them extremely difficult.  The ``VF2${\leftrightarrow}$'' line shows what happens with
VF2 if the graphs are permuted, so that the non-zero labelled instances are given lower vertex
numbers rather than higher vertex numbers. This
makes VF2 behave better, but still extremely poorly compared to the other two solvers; we return to
this in reference to work by \citet{DBLP:conf/edbt/KatsarouNT17} in the following section.
It is also important to note for later that both VF2 variations find some satisfiable and some
unsatisfiable instances extremely hard.

@@

The equation is also used by \citet{DBLP:conf/icde/JiangWYZ07}, who claim that ``usually the
verification time dominates the Query Response Time [s]ince the computational complexity of
$T_{\mathit{iso\_test}}$ is NP-Complete''. They note that \begin{displayquote}``Approximately, the
    value of $T_{\mathit{iso\_test}}$ does not change too much with the difference of query. Thus,
the key to reducing query response time is to minimize the size of the candidate answer
set''.\end{displayquote} This claim becomes understandable when one examines the subgraph matching
algorithm chosen for the verification step \citep{DBLP:journals/jacm/Ullmann76}: as the algorithm
predates techniques like arc-consistent all-different propagation \citep{DBLP:conf/aaai/Regin94}, it
cannot immediately detect unsatisfiability in simple cases like there being two red vertices in the
pattern but only one in the target\footnote{Although interestingly, this algorithm effectively does
forward-checking and has a variable-ordering heuristic, before these concepts appeared in the
constraints literature.}.

>>>>

\begin{abstract}
    The subgraph isomorphism problem involves finding a copy of a pattern graph inside a larger
    target graph. The non-induced version allows extra edges in the target, whilst the induced
    version does not.  Although both subgraph isomorphism variants are \NP-complete, modern
    algorithms can operate comfortably on most real-world graphs with thousands of vertices.
    However, they cannot handle arbitrary instances of this size. We show how to generate ``really
    hard'' random instances for subgraph isomorphism problems, which are computationally challenging
    with only a couple of hundred vertices in the target, and only twenty to thirty pattern
    vertices. For the non-induced version of the problem, these instances lie on a satisfiable /
    unsatisfiable phase transition, whose location we can predict; for the induced variant, much
    richer behaviour is observed, and constrainedness gives a better measure of difficulty than does
    proximity to a phase transition. In both cases, we can use this to understand and improve
    variable and value ordering heuristics.

@@

However, subgraph isomorphism algorithms cannot handle \emph{arbitrary} instances involving hundreds
or thousands of vertices. Experimental evaluations of subgraph isomorphism algorithms are usually
performed using randomly generated graph pairs, sometimes together with a mix of real-world graphs,
and graphs that encode biochemistry and computer vision problems. Using random instances to evaluate
algorithm behaviour can be beneficial, because it provides a way of generating many instances
cheaply, and reduces the risk of over-fitting when tuning design parameters. It also allows us to
control graph parameters such as order, density, or degree distribution, and to study scaling
properties of algorithms with respect to these parameters.

The random instances used in previous evaluations typically come from common datasets
\citep{DBLP:journals/prl/SantoFSV03,DBLP:journals/constraints/ZampelliDS10}, which were generated by
taking a random subgraph of a random (Erd\H{o}s-R\'enyi, scale-free, bounded degree, or mesh) graph
and permuting the vertices. Such instances are guaranteed to be satisfiable---\citet{o:AntonO09}
exploited this property to create large sets of random satisfiable boolean satisfiability instances.
This is the most common approach to generating random subgraph isomorphism instances, meaning
existing benchmark suites contain relatively few non-trivial unsatisfiable instances (although a few
of the patterns in the instances by \citeauthor{DBLP:journals/constraints/ZampelliDS10} have had
extra edges added, making them unsatisfiable). Also, the satisfiable instances tend to be
computationally fairly easy, with most of the difficulty being in dealing with the size of the
model. This has lead to bias in algorithm design, to the extent that some proposed techniques, such
as those of \citet{DBLP:conf/sls/BattitiM07}, will \emph{only} work on satisfiable instances.

The first contribution of this paper is to present and evaluate new methods for creating random
pattern~/ target pairs.  The method we introduce generates both satisfiable and unsatisfiable
instances, and can produce computationally challenging instances with only a few tens of vertices in
the pattern, and 150 vertices in the target. Note that the lack of
unsatisfiable instances for testing purposes cannot be addressed simply by taking a pattern graph
from one of the existing random suites with the ``wrong'' target graph, as this tends to give either
a trivially unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the
case that a relatively small random graph is unlikely to appear in a larger random graph.)

This work builds upon the phase transition phenomena observed in satisfiability and graph colouring
problems first described by \citet{DBLP:conf/ijcai/CheesemanKT91} and
\citet{DBLP:conf/aaai/MitchellSL92}. For subgraph isomorphism in Erd\H{o}s-R\'enyi random graphs, we
identify three relevant control parameters: we can independently alter the edge probability of the
pattern graph, the edge probability of the target graph, and the relative orders (number of
vertices) of the pattern and target graphs.  For non-induced isomorphisms, with the correct choice
of parameters we see results very similar to those observed with boolean satisfiability problems:
there is a phase transition (whose location we can predict) from satisfiable to unsatisfiable, and
we see a solver-independent complexity peak near this phase transition. Additionally, understanding
this behaviour helps us to select better variable- and value-ordering heuristics---this is the
second contribution of this paper.

@@

When labels on vertices are introduced, as is commonly seen in real-world applications and in graph
database systems, richer behaviour emerges, particularly when moving away from a uniform labelling
scheme, and we see that the popular VF2 algorithm behaves much worse than two more recent
constraint-based algorithms in certain cases.  A close look shows that these cases \emph{should} be
easy to solve.

@@

We look at some of the datasets commonly used to test graph database systems, which the literature
suggests are hard to solve. Simple experiments show that the entire perceived difficulty
of each of these datasets comes down to the widespread use of especially poor subgraph isomorphism
algorithms. Our fourth contribution is to show that the use of forward-checking and the right
variable ordering heuristic, even without the more sophisticated recent advances in subgraph
algorithms, makes the entire graph database filter~/ verify paradigm unnecessary.  Finally, we
explain why filter~/ verify (and other recently proposed techniques) cannot be beneficial even on
more challenging instances except if an extremely poor choice of subgraph isomorphism algorithm is
made.

\subsection{Definitions}

Throughout, our graphs are unlabelled, undirected, and do not have any loops.  The \emph{order} of a
graph is the cardinality of its vertex set. We write $\operatorname{V}(G)$ for the vertex set of a
graph $G$. The \emph{complement} of a graph $G$, denoted $\shortoverline{G}$, is the graph with the
same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ if and only if $v$
and $w$ are not adjacent in $G$. We write $G(n, p)$ for an Erd\H{o}s-R\'enyi random graph with $n$
vertices, and an edge between each distinct pair of vertices with independent probability $p$.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. An \emph{induced subgraph
isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$,
then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P \rightarrowtail T$
for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced isomorphism. Observe
that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced isomorphism $i : P
\rightarrowtail T$ which is also a non-induced isomorphism $\shortoverline{i} : \shortoverline{P}
\rightarrowtail \shortoverline{T}$.

\subsection{Experimental Setup}

@@

The Glasgow and LAD solvers store domains, and use backtracking search to build up an assignment of
target vertices (values) to pattern vertices (variables), but differ in terms of inference and
ordering heuristics.  The approach used by VF2 is similar, although the domains of variables are not
stored (in the style of conventional backtracking, rather than forward-checking), and so domain
wipeouts are not detected until an assignment is made. Both the Glasgow and LAD solvers use domain
sizes to guide search, whilst VF2 uses adjacency to previously-assigned vertices.

@@

?? Glasgow not parallel

?? VF3 does not do non-induced

@@

?? LAD and Glasgow not really designed for induced

@@

In the first four rows of \cref{figure:induced} we repeat our experiments, finding induced
isomorphisms. With a pattern of order 10, we get two independent phase transitions: the bottom right
halves of the plots resemble the non-induced results, and the top left halves are close to a mirror
image. The central satisfiable region, which is away from either phase transition, is
computationally easy, but instances near the phase transition are hard.

@@

What about empirical hardness? As before, some instances on the phase transition are hard for all
solvers (although as the number of labels increases, the hardest instances become easier).
Instances far from the phase transition are easy, except for VF2, which occasionally finds
some of the instances with larger numbers of labels very difficult. This behaviour occurs both on
satisfiable instances, and on the flawed unsatisfiable instances deep inside the ``satisfiable''
region.  It is interesting to observe that all such unsatisfiable instances that VF2 finds difficult
have a very small proof of unsatisfiability, with neither Glasgow nor LAD requiring more than one
hundred recursive calls.

@@

We plot the cumulative number of instances solved over time for these instances in
\cref{figure:skewed-cumulative}. Both Glasgow and LAD find all of these instances trivial, but VF2
finds many of them extremely difficult.  The ``VF2${\leftrightarrow}$'' line shows what happens with
VF2 if the graphs are permuted, so that the non-zero labelled vertices are given lower vertex
numbers rather than higher vertex numbers. This
makes VF2 behave better, but still extremely poorly compared to the other two solvers; we return to
this in reference to work by \citet{DBLP:conf/edbt/KatsarouNT17} in the following section.
It is also important to note for later that both VF2 variations find some satisfiable and some
unsatisfiable instances extremely hard.

@@

The equation is also used by \citet{DBLP:conf/icde/JiangWYZ07}, who claim that ``usually the
verification time dominates the Query Response Time [s]ince the computational complexity of
$T_{\mathit{iso\_test}}$ is NP-Complete''. They note that \begin{displayquote}``Approximately, the
    value of $T_{\mathit{iso\_test}}$ does not change too much with the difference of query. Thus,
the key to reducing query response time is to minimize the size of the candidate answer
set''.\end{displayquote} This claim becomes understandable when one examines the subgraph matching
algorithm chosen for the verification step \citep{DBLP:journals/jacm/Ullmann76}: as the algorithm
predates techniques like generalised arc-consistent all-different propagation \citep{DBLP:conf/aaai/Regin94}, it
cannot immediately detect unsatisfiability in simple cases like there being two red vertices in the
pattern but only one in the target\footnote{Although interestingly, this algorithm effectively does
forward-checking and has a variable-ordering heuristic, before these concepts appeared in the
constraints literature.}.
